---
title: "Weight Lifting Method Detection"
author: "Jeremy Voisey"
date: "12 April 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pecent, echo = FALSE}
percent <- function(x) {
    paste0(format(x * 100, digits = 4, nsmall =  2),"%")
}
```

## introduction

In this <a href = "http://groupware.les.inf.puc-rio.br/har">study by Veloso at  al.</a>, accelerometers were placed in various positions on
subjects doing weight lifting exercises. The aim is to identify what method is being used during the exercise.

The sensors were placed on: 

* Upper Arm

* Forearm

* Belt

* Dumbbell

An experienced weight lifter observed the exercise and classified it as:

* A - correct Method

* B - Throwing elbows

* C - Only lifting dumbbell halfway

* D - Only lowering dumbbell halfway

* E - Throwing hips

## Data Processing

### Download

The training and test data (reserved for the final test) is downloaded, if required and loaded.

```{r download}
trainURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainFile <- "pml-training.csv"
if (!file.exists(trainFile)) {
    download.file(trainURL, trainFile)
}
trainingAll <- read.csv(trainFile)

finalTestURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
finalTestFile <- "pml-testing.csv"
if (!file.exists(finalTestFile)) {
    download.file(finalTestURL, finalTestFile)
}
finalTesting <- read.csv(finalTestFile)
```

### Variable Selection

Variables that are missing in the final testing data are removed from both sets.

```{r missingtest}
NAMean <- sapply(finalTesting, function(x){mean(is.na(x))})
notNA <- (NAMean < 0.9)
trainingAll <- trainingAll[,notNA]
finalTesting <- finalTesting[,notNA]
```

The first seven columns contains, weight lifters names, dates etc, are also removed.

```{r nonfeatures}
trainingAll <- trainingAll[,-c(1:7)]
finalTesting <- finalTesting[,-c(1:7)]
```

## Cross Validation

Training data is split into three parts:

* 60% Training

* 20% Base Model Testing and Stack Training (1/2 of 40%)

* 20% Stacked Model Testing


```{r datasplit, warning = FALSE, message = FALSE}
library(caret)
set.seed(271001)
inTrain = createDataPartition(trainingAll$classe, p = 0.6, list = FALSE)
training <- trainingAll[inTrain,]
trainingTMP <- trainingAll[-inTrain,]
inStack = createDataPartition(trainingTMP$classe, p = 1/2, list = FALSE)
trainingStack <- trainingTMP[inStack,]
trainingTest <- trainingTMP[-inStack,]
rm(trainingTMP)
```

## Exporatory Analysis

### featureplot
None of the individual features show huge potential in separating the classe variable.
The six features shown below, being among the best!

```{r features}
featurePlot(x = training[,c("yaw_belt", "accel_belt_z", "magnet_belt_x", "magnet_belt_y", "magnet_arm_y", "magnet_forearm_x")], y = training$classe, labels = c("classe",""))
```

### scatter plot

A scatter plot using just two variables, shows a huge amount of overlap in the
classe variable groups. While there may be a difference in the variability of the classe,
this will not help in identifying individual points. This would suggest a non-linear model
would have most success.

```{r exploratory}
library(ggplot2)
ggplot(training, aes(x = magnet_belt_y, y = magnet_arm_y, colour = classe)) +
    geom_point(alpha = 0.3) + labs(title = "Scatterplot showing very little seperation")
```

## Model Building

Four individual models were tested. Each of these was tested for out of sample
error using the Base model testing and Stack training set.

### Decision Tree

The first model tested, was a simple Decision Tree.

```{r rpart, cache = TRUE, warning = FALSE}
set.seed(43770)
rpartModel <- train(classe ~ ., data = training, method = "rpart")
rpartModel
```

The tree generated is shown below.

```{r treeplot, message = FALSE, warning = FALSE}
library(rattle)
fancyRpartPlot(rpartModel$finalModel)
```

#### Model testing

```{r rpartTest, message = FALSE, warning = FALSE}
rpartPred <- predict(rpartModel, newdata = trainingStack)
rpartCM <- confusionMatrix(rpartPred, trainingStack$classe) #50%
rpartCM
```

Accuracy is very low, at only `r percent(rpartCM$overall[1])`. Note that there is not
even one Prediction of classe = "D".

### Linear Discriminant Analysis

The second model attempted was Linear Discriminant Analysis.

```{r lda, cache = TRUE, warning = FALSE}
set.seed(43770)
ldaModel <- train(classe ~ ., data = training, method = "lda")
ldaModel
```

#### Model testing

```{r ldaTest, message = FALSE, warning = FALSE}
ldaPred <- predict(ldaModel, newdata = trainingStack)
ldaCM <- confusionMatrix(ldaPred, trainingStack$classe) #70%
ldaCM
```

There is some improvement in accuracy, `r percent(ldaCM$overall[1])`

### Stochastic Gradient Boosting

As basic models were not highly successful, more sophisticated models were tested, starting with a boosted model, Stochastic Gradient Boosting.

```{r gbm, cache = TRUE, warning = FALSE}
set.seed(43770)
gbmModel <- train(classe ~ ., data = training, method = "gbm", verbose = FALSE)
gbmModel
```

#### Model testing

```{r gbmTest, message = FALSE, warning = FALSE}
gbmPred <- predict(gbmModel, newdata = trainingStack)
gbmCM <- confusionMatrix(gbmPred, trainingStack$classe) #96%
```

There is a large improvement in accuracy, `r percent(gbmCM$overall[1])`. However there is
still a little room for improvement.

### Random Forest

The final (base) model tried, was Random Forest, as this is known to be very successful as a classification algorithm.

```{r rf, cache = TRUE, warning = FALSE, message = FALSE}
set.seed(43770)
rfModel <- train(classe ~ ., data = training, method = "rf")
rfModel
```

#### Model testing

```{r rfTest, message = FALSE, warning = FALSE}
rfPred <- predict(rfModel, newdata = trainingStack)
rfCM <- confusionMatrix(rfPred, trainingStack$classe) #98.9%
```
This model shows the highest accuracy, `r percent(rfCM$overall[1])`.

This can be seen clearly in a comparision of the confusion matrices below.

```{r confusionplot, message = FALSE, warning = FALSE}
library(dplyr)
normCM1 <- as.data.frame(rpartCM$table / apply(rpartCM$table, 1, sum))
normCM1$Model = 1
normCM2 <- as.data.frame(ldaCM$table / apply(ldaCM$table, 1, sum))
normCM2$Model = 2
normCM3 <- as.data.frame(gbmCM$table / apply(gbmCM$table, 1, sum))
normCM3$Model = 3
normCM4 <- as.data.frame(rfCM$table / apply(rfCM$table, 1, sum))
normCM4$Model = 4
normCM <- bind_rows(normCM1, normCM2, normCM3, normCM4)
modelNames <- factor(1:4, labels = c("Decision Tree", "Linear Discriminant Analysis", "Stochastic Gradient Boosting", "Random Forest"))
normCM$Model <- modelNames[normCM$Model]
normCM$Freq[is.na(normCM$Freq)] <- 0
ggplot(normCM, aes(y = Prediction, x = Reference, fill = Freq)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%1.2f",Freq)), vjust = 1) +
    scale_fill_gradient(low = "yellow", high = "red") +
    scale_y_discrete(limits = c("E", "D", "C", "B", "A")) +
    labs(x = "Actual", title = "Normalized Confusion Matrix") +
    facet_wrap(~Model)
    
```

## Model Stacking

Although the Random Forest Model has very high accuracy, it was decided to see whether
model stacking could improve it further. The predictions from the four models
made on the Base model testing and Stack training set are used as predictors themselves.

The models were stacked using a Random Forest Model.

```{r comb, cache = TRUE, message = FALSE}
stackDF <- data.frame(rpartPred, ldaPred, gbmPred, rfPred, classe = trainingStack$classe)
combMod <- train(classe ~ ., data = stackDF, method = "rf")
```

#### Stacked Model testing
The function stackedPredict obtains the predictions from the four base models
and passes these to the stacked Model.

```{r stackedpredict}
stackedPredict <- function(newdata) {
    rpartPred <- predict(rpartModel, newdata = newdata)
    ldaPred <- predict(ldaModel, newdata = newdata)
    gbmPred <- predict(gbmModel, newdata = newdata)
    rfPred <- predict(rfModel, newdata = newdata)
    stackDF <- data.frame(rpartPred, ldaPred, gbmPred, rfPred)
    predict(combMod, newdata = stackDF)    
}
```


```{r stackedTest}
combPred <- stackedPredict(trainingTest)
stackedCM <- confusionMatrix(combPred, trainingTest$classe) #99.1
```

There is no significant improvement over the Forest Model, with accuracy of `r percent(stackedCM$overall[1])`.


### Final Testing

This was the final model used to make the predictions for the final test set.

As these final test predictions are the answers for a Coursera Quiz,
the honour code prevents me from sharing them, sorry!

However, I will confirm, that they were accurate. Further, they were identical to
the predictions made just using the Random Forest Model.

```{r finalTest, eval= FALSE}
finalPred <- stackedPredict(finalTesting)
data.frame(classe = finalPred)
```

```{r comment, eval = FALSE, echo = FALSE}
#20/20 (B A B A A E D B A A B C B A E E A B B B)
# Stacking unecessary!
identical(combPred,rfPred)
```

#### Citation
The data for this study comes, with thanks, from:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Read more: <a  href = "http://groupware.les.inf.puc-rio.br/har#ixzz4e2aw89ds">http://groupware.les.inf.puc-rio.br/har#ixzz4e2aw89ds</a>